\textit{MSF: Modulated Sub-graph Finder}

### 2018-03-22 MSF_Paper_FA.tex
Corrections

To overcome this limitation, we developed a novel approach to identify
concertedly modulated sub-graphs in the global cell signaling network,
based on the DGEA results of all genes tested. Thereby, expression
patterns of genes are integrated according to the topology of their
interactions and allow potentially to read the flow of information
from the perturbation source to the effectors. The described software,
names \texttt{Modulated Sub-graph Finder} is freely available at
\url{http//:path/to/git}.

@@

To assess the robustness and stability of our method, Poisson
distributed noise was added to the sample read counts \TODO{more details;what samples;what DGEA}. Then DGEA was
carried out on the disturbed data with the same parameters as for the
native data, followed by analysis with \texttt{MSF}. This procedure was carried out 100 times
and every time the genes from the sub-graphs identified from noisy
data were compared to the genes of sub-graphs identified from the
native data. This was also done for the DEG obtained for each run, a
cutoff was used to compare the retrieved DEG from noisy data to the
native data.  The robustness of \texttt{MSF} and the DEG analysis for
the three time-points are shown in~(Fig.~2) \TODO{which time points}. We were very stringent in
adding noise to the counts, which could be seen by the DEG analysis
robustness. For the last time-point (48h~p.i.) 84~\% genes of the
sub-graphs from native data were retrieved, for 24~h~p.i. robustness was
about 79~\%. Robustness was 71~\% at 6~h~p.i., this could be due to that the
infection merely started and the signal transdcution was weak as
compared to the later time-points.

Simulation testing was also conducted on \texttt{MSF}, the objective
was to find less sources than observed in the sub-graphs identified by
\texttt{MSF} from native data. Simulation results showed a chance of
less than 10 percent to get more sources than observed \TODO{we agreed on massively expanding this section or skip it alltogether}.

@@

To demonstrate its usefulness \texttt{MSF} is applied to RNAseq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to
the Filoviridea family; filamentous, enveloped and single stranded RNA
viruses. EBOV causes hemorrhagic fever in humans, inducing the host
innate and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Until now there are no approved antiviral
drugs for the treatment of Ebola virus infection~\cite{Konde,Rhein}.
The initial targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. The
Ebola viral protein VP35 targets the host type~I interferon
IFN-$\alpha / \beta$ to block the early antiviral immune response
~\cite{Prins,Konde,Falasca,Cardenas,Olejnik}. It has been proposed
that IFN-$\alpha / \beta$ should be tested against Ebola for its
antiviral activity through clinical trials~\cite{Konde}. The aim of
testing the EBOV infection data with \texttt{MSF} was to identify the
modulated sub-graphs and check if \texttt{MSF} is able to identify the
IFN-$\alpha / \beta$ gene as one of the sources. \TODO{why this data set?}

EBOV infection count data was downloaded from GEO
(GSE84188). Differential gene expression analysis was performed on the
count data with edgeR package (version 3.4.2)~\cite{edgeR}. The DEG
list generated by edgeR was used as an input for \texttt{MSF} \TODO{above you said MSF does not use a DEG gene list}. The
EBOV infection experiments describe the course of infection at three time-points 6, 24 and 48 hour post
infection (h~p.i.). For the earliest time point at 6 h~p.i., five modulated sub-graphs were
identified \TODO{with how many genes each?}. Most of the genes part of the sub-graphs were chemokines
(CXCL10, CCL8) and Interleukin genes (IL6, IL27, IL23) which serve as
a subset of CD4+ T helper cells. IFNB1 and IFNA1 were both identified
as two of the possible sources. Most of the sources identified by
\texttt{MSF} were type~I interferon induced genes. At 24h~p.i. nine
modulated sub-graphs were identified, IFNA1 was identified as one of
the sources. For the last time-point again IFNB1 and IFNA1 were
identified as the two sources out of the possible sources from seven
modulated sub-graphs identified.

@@

\subsection*{Modulated sub-graphs at 6h~p.i.}

 Three main modulated sub-graphs identified by \texttt{MSF} at 6h~p.i.
 are shown in~(Fig.~3). The graph shows different colored nodes as
 genes associated to KEGG pathways and how they cross talk. The KEGG
 association and visualization on the modulated sub-graphs was done
 using StringApp~\cite{StringApp} in Cytoscape~\cite{Cyto}. The graph
 unveils how more than one stimulus coincides to prevent the
 progression of infection. Sub-graph 1 shows toll-like receptor,
 cytokine, chemokine and jak-stat genes leading to TNF and eventually
 going into apoptosis. Sub-graph 2 shows Extra-cellular matrix (ECM)
 receptor interaction genes which are suggested to be in interaction
 with Ebola glycoprotein (GP)~\cite{Veljkovic}. ECM receptor, Focal
 adhesion and P13k-Akt signaling genes are the sources from one end of
 the sub-graph. P13k-Akt signaling goes to FoxO signaling which
 eventually leads to cell cycle, which at later time-points is
 disrupted. From the other end of sub-graph 2 toll-like recpetor
 combined with cytosolic DNA sensing and cytokine-chemokine signaling
 activates Jak-stat which leads to the start of apoptosis with a
 activation of a few apoptosis genes. Cytokine is also seen to sink
 into natural killer mediated cytotoxicity. Sub-graph 3 shows how some
 of the cytokine and chemokine genes activate the natural killer
 mediated cytotoxicity and apoptosis.

@@

To further understand the functional role of the sub-graphs identified
by \texttt{MSF}, we linked the sub-graphs to the predefined KEGG
pathways and Reactome pathways through gene enrichment analysis. We
used Cytoscape~\cite{Cyto} plugins to identify the enriched patwhays,
CytoKegg~\cite{Cytokegg} for KEGG and Reactome FI~\cite{Reactome} for
Reactome. We compared the pathways enriched in \texttt{MSF} identified
sub-graphs to pathway analysis by SPIA~\cite{Tarca} for KEGG and
Reactome pathway enrichment analysis for Reactome.

At 6h~p.i. Cytokegg identified the sub-graph genes to be enriched in
Toll-like receptor signaling, TNF signaling, IL-17 signaling and
NF-kappa signaling pathways, as also stated by~\cite{Olejnik}. The
earliest response EBOV induces on cytokine is via TLR4-mediated
signaling~\cite{Olejnik}, Gene enrichment analysis of sub-graph genes
at 6h~p.i. showed toll-like receptor signaling being most significantly
dis-regulated pathway, this important signaling pathway was not
identified by SPIA along with RIG-I-like receptor signaling pathway,
TNF signaling pathway and IL-17 signaling (Table 1 in the supplement
material). On the later time-points \texttt{MSF} and SPIA showed
consensus on the important pathways e.g Chemokine signaling,
Cytokine-Cytokine receptor interaction, NF-kappa B signaling,
Cytosolic DNA-sensing and RIG-I-like receptor signaling as being
dis-regulated (Table 2-3 in the supplement).

Likewise we compared gene enrichment results of sub-graph genes
identified by \texttt{MSF} using Reactome FI to the Reactome
predefined pathway enrichment analysis. Although \texttt{MSF} and
Reactome showed good agreement on the dis-regulated pathways e.g
Signaling by Interleukins, Interleukin-10 signaling, Cytokine
Signaling in Immune system, Chemokine receptors bind chemokines,
RIG-I/MDA5 mediated induction of IFN-alpha/beta pathways, NF-kB
activation, Reactome as well could not identify Toll-like receptor
signaling (Table 1-3 in the supplement material) . This shows that
\texttt{MSF} is able to identify the important modulated genes in the
sub-graphs at the earliest time-point of infection.

@@

\begin{figure*}[p]
	\centering
	\fbox{
		\begin{minipage}{17 cm}
			\includegraphics[width=16cm]{DEGvsMSF2.png}
			\caption{Graph shows the percentage of DEG
                          analysis genes and MSF identified sub-graph
                          genes recall rates 100 times for the three
                          different time points after the addition of
                          Poisson distributed noise.}
			\label{label}
		\end{minipage}
	}
\end{figure*}

\begin{figure*}
	\centering
	\fbox{
		\begin{minipage}{17 cm}
	\includegraphics[width=1.05\linewidth]{SeperatedMergedNetwork.png}
	\caption{Shows the three modulated sub-graphs identified by
          MSF at 6h~p.i., where nodes are genes and edges are the
          interactions. The node coloring is associated to KEGG
          pathways referring to the colors in the legend. The graph
          edges are from Reactome. Important genes to EBOV infection
          as from literature are enlarged in the graph. The signal
          flow corresponding to each sub-graph is depicted as simple
          flow chart against each sub-graph.}
		\end{minipage}
}
\end{figure*}

>>>>

To overcome this limitation, we developed a novel approach to identify
concertedly modulated sub-graphs in the global cell signaling network,
based on the DGEA results of all genes tested. Thereby, expression
patterns of genes are integrated according to the topology of their
interactions and allow potentially to read the flow of information
from the perturbation source to the effectors. The described software,
names \texttt{Modulated Sub-graph Finder} is freely available at
\url{https://github.com/MariamFarman/Modulated-SubGraph-Finder}.

@@

@@

To demonstrate its usefulness \texttt{MSF} is applied to RNAseq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to
the Filoviridea family; filamentous, enveloped and single stranded RNA
viruses. EBOV causes hemorrhagic fever in humans, inducing the host
innate and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Until now there are no approved antiviral
drugs for the treatment of Ebola virus infection~\cite{Konde,Rhein}.
The initial targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. The
Ebola viral protein VP35 targets the host type~I interferon
IFN-$\alpha / \beta$ to block the early antiviral immune response
~\cite{Prins,Konde,Falasca,Cardenas,Olejnik}. It has been proposed
that IFN-$\alpha / \beta$ should be tested against Ebola for its
antiviral activity through clinical trials~\cite{Konde}. This data was chosen for analysis because it was the most recent EBOV infection data and the infection was carried out in human macrophages, which would help see the innate immune response of host. The aim of
testing this data with \texttt{MSF} was to identify the
modulated sub-graphs and check if \texttt{MSF} is able to identify the
IFN-$\alpha / \beta$ gene as one of the sources.

EBOV infection count data was downloaded from GEO
(GSE84188). Differential gene expression analysis was performed on the
count data with edgeR package (version 3.4.2)~\cite{edgeR}. The DEG
analysis results generated by edgeR were used as input for \texttt{MSF}. The
EBOV infection experiments describe the course of infection at three time-points 6, 24 and 48 hour post
infection (hpi). For the earliest time point at 6 hpi, five modulated sub-graphs were
identified with 41, 107, and 18 genes. Two of the five sub-graphs were less than 4 genes long. Most of the genes part of the sub-graphs were cytokines, chemokines
(CXCL10, CCL8) and Interleukin genes (IL6, IL27, IL23) which serve as
a subset of CD4+ T helper cells. IFNB1 and IFNA1 were both identified
as two of the possible sources. Most of the sources identified by
\texttt{MSF} were type~I interferon induced genes. At 24 hpi nine
modulated sub-graphs were identified, IFNA1 was identified as one of
the sources. For the last time-point again IFNB1 and IFNA1 were
identified as the two sources out of the possible sources from seven
modulated sub-graphs identified.

@@

To assess the robustness and stability of our method, Poisson
distributed noise was added to the read counts at the 3 time-points. Then DGEA was
carried out on the disturbed data with the same parameters as for the
native data using edgeR, followed by analysis with \texttt{MSF}. This procedure was carried out 100 times
and every time the genes from the sub-graphs identified from noisy
data were compared to the genes of sub-graphs identified from the
native data. This was also done for the DEG obtained for each run, a
cutoff was used to compare the retrieved DEG from noisy data to the
native data.  The robustness of \texttt{MSF} and the DEG analysis for
the time-point 6, 24, and 48 hpi are shown in Fig.~\ref{fig:DEGvsMSF}. We were very stringent in
adding noise to the counts, which could be seen by the DEG analysis
robustness. For the last time-point (48h~p.i.) 84~\% genes of the
sub-graphs from native data were retrieved, for 24~h~p.i. robustness was
about 79~\%. Robustness was 71~\% at 6~h~p.i., this could be due to that the
infection merely started and the signal transdcution was weak as
compared to the later time-points.

Simulation testing was also conducted on \texttt{MSF}. For each time-point, one \texttt{MSF} identified sub-graph was taken. A random graph was generated with the same number of genes as the sub-graphs and the predicted sources were counted. The objective
was to find less sources in the simulated sub-graphs than observed in the real sub-graphs identified by
\texttt{MSF}. Simulation results showed a chance of $<$10$\%$
of the 1000 random sub-graphs generated to get more sources than observed.

\subsection*{Modulated sub-graphs at 6 hpi}

 Three main modulated sub-graphs identified by \texttt{MSF} at 6 hpi
 are shown in Fig.~\ref{fig:Sub-graph6hpi}. The graph shows different colored nodes as
 genes associated to different KEGG pathways and how they cross talk. The KEGG
 association and visualization on the modulated sub-graphs was done
 using StringApp~\cite{StringApp} in Cytoscape~\cite{Cyto}. The graph
 unveils how more than one stimulus coincides to prevent the
 progression of infection. Sub-graph 1 shows toll-like receptor,
 cytokine, chemokine and jak-stat genes leading to TNF and eventually
 going into apoptosis. Sub-graph 2 shows Extra-cellular matrix (ECM)
 receptor interaction genes which are suggested to be in interaction
 with Ebola glycoprotein (GP)~\cite{Veljkovic}. ECM receptor, Focal
 adhesion and P13k-Akt signaling genes are the sources from one end of
 the sub-graph. P13k-Akt signaling goes to FoxO signaling which
 eventually leads to cell cycle, which at later time-points is
 disrupted. From the other end of sub-graph 2 toll-like recpetor
 combined with cytosolic DNA sensing and cytokine-chemokine signaling
 activates Jak-stat which leads to the start of apoptosis with a
 activation of a few apoptosis genes. Cytokine is also seen to sink
 into natural killer mediated cytotoxicity. Sub-graph 3 shows how some
 of the cytokine and chemokine genes activate the natural killer
 mediated cytotoxicity and apoptosis.

@@

To further understand the functional role of the sub-graphs identified
by \texttt{MSF}, we linked the sub-graphs to the predefined KEGG
pathways and Reactome pathways through gene enrichment analysis. We
used Cytoscape~\cite{Cyto} plugins to identify the enriched patwhays,
CytoKegg~\cite{Cytokegg} for KEGG and Reactome FI~\cite{Reactome} for
Reactome. A comparison to other tools was performed on the same dataset using each software's default parameters. We compared the pathways enriched in \texttt{MSF} identified
sub-graphs to pathway analysis by SPIA~\cite{Tarca} for KEGG and
Reactome pathway enrichment analysis for Reactome.

At 6 hpi Cytokegg identified the sub-graph genes to be enriched in
Toll-like receptor signaling, TNF signaling, IL-17 signaling and
NF-kappa signaling pathways, as also stated by~\cite{Olejnik}. The
earliest response EBOV induces on cytokine is via TLR4-mediated
signaling~\cite{Olejnik}, Gene enrichment analysis of sub-graph genes
at 6 hpi showed toll-like receptor signaling being most significantly
dis-regulated pathway, this important signaling pathway was not
identified by SPIA along with RIG-I-like receptor signaling pathway,
TNF signaling pathway and IL-17 signaling. On the later time-points \texttt{MSF} and SPIA showed
consensus on the important pathways e.g Chemokine signaling,
Cytokine-Cytokine receptor interaction, NF-kappa B signaling,
Cytosolic DNA-sensing and RIG-I-like receptor signaling as being
dis-regulated.

Likewise we compared gene enrichment results of sub-graph genes
identified by \texttt{MSF} using Reactome FI to the Reactome
predefined pathway enrichment analysis. Although \texttt{MSF} and
Reactome showed good agreement on the dis-regulated pathways e.g
Signaling by Interleukins, Interleukin-10 signaling, Cytokine
Signaling in Immune system, Chemokine receptors bind chemokines,
RIG-I/MDA5 mediated induction of IFN-alpha/beta pathways, NF-kB
activation, Reactome as well could not identify Toll-like receptor
signaling. This shows that
\texttt{MSF} is able to identify the important modulated genes in the
sub-graphs at the earliest time-point of infection. A summary of the comparison between the different tools is given in the supplement Table 1.

@@

\begin{figure*}[p]
	\centering
	\fbox{
		\begin{minipage}{17 cm}
			\includegraphics[width=16cm]{DEGvsMSF2.png}
			\caption{Graph shows the percentage of DEG
                          analysis genes and MSF identified sub-graph
                          genes recall rates 100 times for the three
                          different time points after the addition of
                          Poisson distributed noise.\label{fig:DEGvsMSF}}
			\label{label}
		\end{minipage}
	}
\end{figure*}

\begin{figure*}
	\centering
	\fbox{
		\begin{minipage}{17 cm}
	\includegraphics[width=1.05\linewidth]{SeperatedMergedNetwork.png}
	\caption{Shows the three modulated sub-graphs identified by
          MSF at 6 hpi, where nodes are genes and edges are the
          interactions. The node coloring is associated to KEGG
          pathways referring to the colors in the legend. The graph
          edges are from Reactome. Important genes to EBOV infection
          as from literature are enlarged in the graph. The signal
          flow corresponding to each sub-graph is depicted as simple
          flow chart against each sub-graph.\label{fig:Sub-graph6hpi}}
		\end{minipage}
}
\end{figure*}
